---
title: "Differential Expression Analysis with DESeq2"
author: biobytes
output: 
  html_document:
    code_folding: hide
    df_print: paged
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(GenomicFeatures)
library(edgeR)
library(grid)
library(org.Hs.eg.db)
library(DESeq2)
library(clusterProfiler)
library(tidyverse)
library(readxl)
library(RColorBrewer)
library(pheatmap)

```

# Load data 

```{r}
setwd("/Users/isabellacotta/Library/CloudStorage/GoogleDrive-icottagalvao@gmail.com/My Drive/Doutorado/Disciplinas/MO413 - Ciência de dados e visualização em saúde/ProjetoFinal")

# Read in count matrix 
counts <- read.delim("data/rsem.merged.gene_counts.tsv") %>% 
  mutate(gene_id = str_remove(gene_id, "\\..+")) %>% 
  select(-transcript_id.s.) %>%
  filter(!duplicated(gene_id)) %>%
  column_to_rownames("gene_id")

# Map ENSEMBL into symbols
symb <- clusterProfiler::bitr(rownames(counts), 
                              fromType = "ENSEMBL", toType = "SYMBOL", 
                              OrgDb = org.Hs.eg.db)
counts <- counts[symb$ENSEMBL, ]


# Read in metadata
meta <- read_excel("data/std_meta_Aus.xlsx")

```


# Tidy data

```{r, warning = FALSE}
# Set same names for sample names in the metadata and the count matrix 
counts <- counts[meta$sample] # same order

meta %>% dplyr::count(diagnosis)
rownames(meta) <- meta$sample
counts <- counts[meta$sample]
#all(colnames(counts) == rownames(meta))
```


# Create DESeq2 Object 

```{r}
# Create DESeq2 object 
meta <- meta %>% mutate(diagnosis = str_remove(diagnosis, " |-"),
                        diagnosis = factor(diagnosis), 
                        lobe = factor(lobe))
dds <- DESeqDataSetFromMatrix(countData = round(counts), 
                              colData = meta, 
                              design = ~ diagnosis + lobe)

# Prefilter
keep <- rowSums(counts(dds) >= 10) >= 5
dds <- dds[keep,]
```


# Differential Expression Test

The design was specified when creating the object. 

```{r}
dds <- DESeq(dds)
```

Extract normalized/transformed counts:

```{r}
setwd("~/Library/CloudStorage/GoogleDrive-icottagalvao@gmail.com/My Drive/Doutorado/Disciplinas/MO413 - Ciência de dados e visualização em saúde/ProjetoFinal")

norm <- counts(dds, normalized = TRUE)

# Export normalized counts 
write.csv(norm, "results/normalized_counts_DESeq2.csv", row.names = TRUE)

# Transformed variance: - for visualization
vsd <- vst(dds, blind = FALSE)
```



# Results 

## Sample distances 

```{r}
# Get sample-to-sample distances 
dist <- dist(t(assay(vsd)))

# Distance matrix 
sampleDistMatrix <- as.matrix( dist )
#colnames(sampleDistMatrix) <- NULL

colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

annotation_rows = data.frame(
    diagnosis = vsd$diagnosis
)
rownames(annotation_rows) <- row.names(sampleDistMatrix)

pal <- brewer.pal(5, "Set2")
names(pal) <- levels(vsd$diagnosis)

ann_colors = list(
  diagnosis = pal
)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=dist,
         clustering_distance_cols=dist,
         annotation_row = NULL,
         annotation_col = annotation_rows,
         show_colnames = F,
         show_rownames = F,
         annotation_colors = ann_colors,
         col=colors,
         clustering_method = "complete"
) 
```
```{r}
vsd.sub <- vsd[, vsd$diagnosis %in% c("Control", "FCDIIb")]

# Get sample-to-sample distances 
dist <- dist(t(assay(vsd.sub)))

# Distance matrix 
sampleDistMatrix <- as.matrix( dist )
#colnames(sampleDistMatrix) <- NULL

colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

annotation_rows = data.frame(
    Diagnosis = vsd.sub$diagnosis
)
rownames(annotation_rows) <- row.names(sampleDistMatrix)

pal <- brewer.pal(5, "Set2")
names(pal) <- levels(vsd$diagnosis)

ann_colors = list(
  Diagnosis = c("Control" = "darkgreen", "FCDIIb" = "orange")
)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=dist,
         clustering_distance_cols=dist,
         annotation_row = NULL,
         annotation_col = annotation_rows,
         show_colnames = F,
         show_rownames = F,
         annotation_colors = ann_colors,
         col=colors,
         clustering_method = "complete"
) 

# Export as svg 
ggsave("results/sample_distance.svg",
       width = 6, 
       height = 4, 
       dpi = 300)
```


## FCD IIb vs. Non-Epileptic Controls

```{r} 
res <- results(dds, contrast = c("diagnosis", "FCDIIb", "Control"))
```

All genes: 

```{r}
summary(res)
#resultsNames(dds)

res <- res %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>% 
  left_join(symb, by = c("gene" = "ENSEMBL")) %>% 
  select(gene, SYMBOL, log2FoldChange, pvalue, padj) %>% 
  mutate(across(c("log2FoldChange", padj), round, digits = 4))
DT::datatable(res, options = list(ordering = FALSE), filter = "top")
```

Significant changes (FC > 1.5):

```{r}
sig <- res %>% 
  filter(padj < 0.1, abs(log2FoldChange) > log2(1.5)) %>% 
  arrange(desc(log2FoldChange))

sum(sig$log2FoldChange > 0)

DT::datatable(sig, options = list(ordering = FALSE), filter = "top")
```

Export list:

```{r}
setwd("~/Library/CloudStorage/GoogleDrive-icottagalvao@gmail.com/My Drive/Doutorado/Disciplinas/MO413 - Ciência de dados e visualização em saúde/ProjetoFinal")

# All genes with fold changes 
write.csv(res, "results/all_DEGs_FCDIIbvsControl.csv", row.names = FALSE)

# Significant genes 
write.csv(sig, "results/significant_DEGs_FCDIIbvsControl.csv", row.names = FALSE)

```


### Visualization 

Volcano plot:

```{r}
# Create new column if up/down regulated 
res <- res %>% 
  mutate(Status = case_when(log2FoldChange > log2(1.5) & padj < 0.1 ~ "Up-regulated", log2FoldChange < -log2(1.5) & padj < 0.1 ~ "Down-regulated", TRUE ~ "Not significant"),
         label = ifelse((padj < 0.1 & abs(log2FoldChange) > 2.5), SYMBOL, "")) 

ggplot(res, aes(x = log2FoldChange, y = -log10(padj), color = Status, label = label)) + 
  geom_vline(xintercept = c(-log2(1.5), log2(1.5)), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.1), col = "gray", linetype = 'dashed') + 
  geom_point() + 
  ggrepel::geom_text_repel(size = 2, color = "black")+
  ylim(0, 5) + 
  scale_color_manual(values = c("Up-regulated"="red", "Down-regulated"="blue", "Not significant" = "black")) + 
  theme_classic() 


# Export as svg 
ggsave("results/volcano_plot.svg",
       width = 6, 
       height = 4, 
       dpi = 300)
```




### Functional Enrichment 

We'll perform Gene Set Enrichment Analysis on Gene Ontology terms. 

```{r}
fc <- res$log2FoldChange
names(fc) <- res$SYMBOL
fc <- sort(fc, decreasing = TRUE)
  
ego <- gseGO(geneList = fc, 
             OrgDb = org.Hs.eg.db,
             ont = "BP", 
             keyType = "SYMBOL",
             eps = 1e-20)
```

All functions:

```{r}
# All functions
funs <- ego@result %>% 
  as.data.frame %>% 
  select(Description, NES, core_enrichment) %>% 
  arrange(desc(NES))
DT::datatable(funs, options = list(ordering = FALSE), filter = "top")
```

Simplified:

```{r}
ego.simp <- clusterProfiler::simplify(ego, cutoff = 0.4)
funs.simp <- ego.simp@result %>% 
  as.data.frame %>% 
  select(Description, NES, core_enrichment) %>% 
  arrange(desc(NES))
DT::datatable(funs.simp, options = list(ordering = FALSE))
```


```{r}
# Export simplified functions table 
write.csv(funs.simp, 
          "results/GSEA.simp.csv")
```

# Reproducibility 

This document was last rendered on: `r Sys.time()`.

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>